---
title: "From Distance to Decision: A Study of K-Means and KNN"
author: "Srujith Kapuluri"
date: 06-11-2025
---

<!-- _todo: do two analyses.  Do one of either 1a or 1b, AND one of either 2a or 2b._ -->

## 1a. K-Means

<!-- _todo: write your own code to implement the k-means algorithm.  Make plots of the various steps the algorithm takes so you can "see" the algorithm working.  Test your algorithm on the Palmer Penguins dataset, specifically using the bill length and flipper length variables.  Compare your results to the built-in `kmeans` function in R or Python._ -->


```{r}
#| message: false
#| warning: false
#| echo: true
#| code-fold: true
#| code-summary: "Load Libraries"
library(tidyverse)
library(palmerpenguins)
penguins_data <- penguins %>%
  select(bill_length_mm, flipper_length_mm) %>%
  drop_na()

X <- as.matrix(penguins_data)
```

```{r}
#| message: false
#| warning: false
#| echo: true
#| code-fold: true
#| code-summary: "Kmeans functions"
# Helper: Euclidean distance
euclidean <- function(a, b) {
  sqrt(sum((a - b)^2))
}

assign_clusters <- function(X, centroids) {
  apply(X, 1, function(point) {
    dists <- apply(centroids, 1, function(centroid) euclidean(point, centroid))
    which.min(dists)
  })
}

# Recalculate centroids
update_centroids <- function(X, clusters, k) {
  sapply(1:k, function(i) colMeans(X[clusters == i, , drop = FALSE])) %>% t()
}
```
```{r}
#| label: kmeans-animation
#| echo: true
#| code-fold: true
#| code-summary: "K-Means clustering as a GIF"
#| message: false
#| warning: false
#| results: "markup"

library(tidyverse)
library(gganimate)
library(gifski)

# Set seed and parameters
set.seed(123)
k <- 3
n_iter <- 5
init_idx <- sample(1:nrow(X), k)
centroids <- X[init_idx, ]

# Build dataset with cluster assignments for each iteration
all_iters <- list()

for (i in 1:n_iter) {
  clusters <- assign_clusters(X, centroids)
  df <- as.data.frame(X)
  df$cluster <- factor(clusters)
  df$iter <- i
  centroids <- update_centroids(X, clusters, k)
  all_iters[[i]] <- df
}

anim_df <- bind_rows(all_iters)

# Build animation with bold + centered title
ggplot(anim_df, aes(x = bill_length_mm, y = flipper_length_mm, color = cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(title = "K-Means Iteration: {closest_state}",
       x = "Bill Length (mm)",
       y = "Flipper Length (mm)") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14)
  ) +
  transition_states(iter, transition_length = 1, state_length = 1) +
  ease_aes('linear')
```



```{r}
#| label: builtin-kmeans
#| echo: true
#| code-fold: true
#| code-summary: "Built-in KMeans Plot"
#| message: false
#| warning: false
#| results: "markup"

kmeans_builtin <- kmeans(X, centers = 3, nstart = 10)

df_builtin <- data.frame(X, cluster = factor(kmeans_builtin$cluster))
centroids_builtin <- as.data.frame(kmeans_builtin$centers)
centroids_builtin$cluster <- factor(1:k)

ggplot(df_builtin, aes(x = bill_length_mm, y = flipper_length_mm, color = cluster)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_point(data = centroids_builtin, aes(x = bill_length_mm, y = flipper_length_mm),
             color = "black", size = 4, shape = 8) +
  labs(title = "Built-in KMeans Result",
       x = "Bill Length (mm)", y = "Flipper Length (mm)") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14)
  )
```

::: {.callout-note title="Interpretation: Built-in KMeans Clustering Result"}

This plot displays the output of R's built-in `kmeans()` function applied to the penguins dataset, using `bill_length_mm` and `flipper_length_mm` as clustering variables with `k = 3`.

- Each point is colored based on its assigned cluster.
- Centroids are shown as black asterisk markers.
- The clusters appear well-separated in 2D space, illustrating how `kmeans()` partitions data based on Euclidean distance.

This plot serves as a benchmark to validate the results of the custom K-Means implementation. The similarity of cluster shapes and centroids between this and the manual version demonstrates that the algorithm was correctly implemented.
:::


<!-- _todo: Calculate both the within-cluster-sum-of-squares and silhouette scores (you can use built-in functions to do so) and plot the results for various numbers of clusters (ie, K=2,3,...,7). What is the "right" number of clusters as suggested by these two metrics?_ -->
```{r}
#| label: cluster-metrics
#| echo: true
#| message: false
#| warning: false
#| results: "markup"
#| code-fold: true
#| code-summary: "Elbow & Silhouette Scores for K=2 to 7"

library(cluster)
library(ggplot2)
library(gridExtra)

# Compute WCSS and silhouette
wcss <- numeric()
silhouette_avg <- numeric()
k_values <- 2:7

for (k in k_values) {
  km <- kmeans(X, centers = k, nstart = 10)
  wcss[k - 1] <- km$tot.withinss
  sil <- silhouette(km$cluster, dist(X))
  silhouette_avg[k - 1] <- mean(sil[, 3])
}

df_metrics <- data.frame(
  k = k_values,
  WCSS = wcss,
  Silhouette = silhouette_avg
)

# Plot 1: Elbow
p1 <- ggplot(df_metrics, aes(x = k, y = WCSS)) +
  geom_line(color = "#5B8FA8", linewidth = 1) +
  geom_point(size = 2, color = "#5B8FA8") +
  labs(title = "Elbow Method", x = "K", y = "WCSS") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.title = element_text(size = 12),
    panel.grid.major = element_line(color = "#DADADA")
  )

# Plot 2: Silhouette
p2 <- ggplot(df_metrics, aes(x = k, y = Silhouette)) +
  geom_line(color = "#5B8FA8", linewidth = 1) +
  geom_point(size = 2, color = "#5B8FA8") +
  labs(title = "Silhouette Score", x = "K", y = "Avg Silhouette") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.title = element_text(size = 12),
    panel.grid.major = element_line(color = "#DADADA")
  )

# Arrange side by side
grid.arrange(p1, p2, ncol = 2)
```

::: {.callout-note title="Interpretation: Elbow & Silhouette Scores"}


To figure out the best number of clusters for K-Means, I used two visuals side by side that each tell a different part of the story.
Elbow Method (Left Plot)

This one tracks how the within-cluster sum of squares (WCSS) drops as we increase K from 2 to 7. Naturally, WCSS always goes down—more clusters mean less distance within them. But the real question is: when does adding more clusters stop giving real value?

In this case, you can see that the drop levels off around K = 3 or 4—that’s the "elbow." Past that, we’re getting smaller gains and possibly just splitting natural groupings.

Silhouette Score (Right Plot)

This one checks how cleanly separated the clusters are. It scores each K based on how close points are to their own cluster versus others. Higher = better.

The standout? K = 3 again—it had the highest average silhouette width, meaning it balanced tight clusters and good separation.
:::



<!-- _If you want a challenge, add your plots as an animated gif on your website so that the result looks something like [this](https://www.youtube.com/shorts/XCsoWZU9oN8)._ -->







## 2a. K Nearest Neighbors

<!-- _todo: use the following code (or the python equivalent) to generate a synthetic dataset for the k-nearest neighbors algorithm.  The code generates a dataset with two features, `x1` and `x2`, and a binary outcome variable `y` that is determined by whether `x2` is above or below a wiggly boundary defined by a sin function._ -->

```{r}
# gen data -----
set.seed(42)
n <- 100
x1 <- runif(n, -3, 3)
x2 <- runif(n, -3, 3)
x <- cbind(x1, x2)

# define a wiggly boundary
boundary <- sin(4*x1) + x1
y <- ifelse(x2 > boundary, 1, 0) |> as.factor()
dat <- data.frame(x1 = x1, x2 = x2, y = y)
```

```{r}
#| label: gen-knn-data
#| echo: true
#| code-fold: true
#| code-summary: "Generate synthetic data for KNN"
#| message: false
#| warning: false
#| results: "markup"

# Generate data
set.seed(42)
n <- 100
x1 <- runif(n, -3, 3)
x2 <- runif(n, -3, 3)
x <- cbind(x1, x2)

# Define boundary
boundary <- sin(4 * x1) + x1
y <- ifelse(x2 > boundary, 1, 0) |> as.factor()

# Combine into data frame
dat <- data.frame(x1 = x1, x2 = x2, y = y)
```



<!-- _todo: plot the data where the horizontal axis is `x1`, the vertical axis is `x2`, and the points are colored by the value of `y`.  You may optionally draw the wiggly boundary._ -->

```{r}
#| label: plot-knn-synthetic
#| echo: true
#| code-fold: true
#| code-summary: "Plot x1 vs x2 colored by y"
#| message: false
#| warning: false
#| results: "markup"

library(ggplot2)

ggplot(dat, aes(x = x1, y = x2, color = y)) +
  geom_point(size = 2, alpha = 0.8) +
  stat_function(fun = function(x) sin(4 * x) + x,
                color = "black", linetype = "dashed", linewidth = 1) +
  labs(title = "Synthetic Dataset for KNN",
       subtitle = "Dashed line = Decision Boundary (sin(4x1) + x1)",
       x = "x1", y = "x2", color = "y") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5)
  )
```



<!-- _todo: generate a test dataset with 100 points, using the same code as above but with a different seed._ -->

```{r}
#| label: gen-knn-test
#| echo: true
#| code-fold: true
#| code-summary: "Generate test dataset (new seed)"
#| message: false
#| warning: false
#| results: "markup"

# Generate test data with different seed
set.seed(99)
n_test <- 100
x1_test <- runif(n_test, -3, 3)
x2_test <- runif(n_test, -3, 3)
boundary_test <- sin(4 * x1_test) + x1_test
y_test <- ifelse(x2_test > boundary_test, 1, 0) |> as.factor()

# Combine into test data frame
test_data <- data.frame(x1 = x1_test, x2 = x2_test, y = y_test)

# Preview
head(test_data)
```


<!-- _todo: implement KNN by hand.  Check you work with a built-in function -- eg, `class::knn()` or `caret::train(method="knn")` in R, or scikit-learn's `KNeighborsClassifier` in Python._ -->

```{r}
#| label: knn-manual
#| echo: true
#| code-fold: true
#| code-summary: "Manual KNN implementation (k = 5)"
#| message: false
#| warning: false
#| results: "markup"

# Manual KNN function
euclidean_dist <- function(x1, x2) {
  sqrt(sum((x1 - x2)^2))
}

knn_predict <- function(train_X, train_y, test_X, k = 5) {
  preds <- sapply(1:nrow(test_X), function(i) {
    dists <- apply(train_X, 1, function(x) euclidean_dist(x, test_X[i, ]))
    neighbors <- order(dists)[1:k]
    majority <- names(sort(table(train_y[neighbors]), decreasing = TRUE))[1]
    return(majority)
  })
  factor(preds, levels = levels(train_y))
}

# Prepare feature matrices
train_X <- dat[, c("x1", "x2")]
train_y <- dat$y
test_X <- test_data[, c("x1", "x2")]
test_y <- test_data$y

# Predict manually
manual_preds <- knn_predict(train_X, train_y, test_X, k = 5)

# Accuracy
manual_acc <- mean(manual_preds == test_y)
manual_acc
```

```{r}
#| label: knn-built-in
#| echo: true
#| code-fold: true
#| code-summary: "Check with class::knn()"
#| message: false
#| warning: false
#| results: "markup"

library(class)

# Built-in knn prediction
builtin_preds <- knn(train = train_X,
                     test = test_X,
                     cl = train_y,
                     k = 5)

# Accuracy
builtin_acc <- mean(builtin_preds == test_y)
builtin_acc
```


<!-- _todo: run your function for k=1,...,k=30, each time noting the percentage of correctly-classified points from the test dataset. Plot the results, where the horizontal axis is 1-30 and the vertical axis is the percentage of correctly-classified points. What is the optimal value of k as suggested by your plot?_  -->

```{r}
#| label: knn-accuracy-curve
#| echo: true
#| code-fold: true
#| code-summary: "Test accuracy across k = 1 to 30"
#| message: false
#| warning: false
#| results: "markup"

# Accuracy for each k
k_vals <- 1:30
acc_vals <- numeric(length(k_vals))

for (i in seq_along(k_vals)) {
  k <- k_vals[i]
  preds_k <- knn_predict(train_X, train_y, test_X, k)
  acc_vals[i] <- mean(preds_k == test_y)
}

# Create dataframe
knn_perf <- data.frame(k = k_vals, accuracy = acc_vals * 100)

# Plot with subtle color styling
library(ggplot2)

ggplot(knn_perf, aes(x = k, y = accuracy)) +
  geom_line(color = "#5B8FA8", linewidth = 1) +
  geom_point(color = "#5B8FA8", size = 2) +
  labs(title = "KNN Test Accuracy vs. Number of Neighbors",
       x = "Number of Neighbors (k)",
       y = "Test Accuracy (%)") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(size = 12),
    panel.grid.major = element_line(color = "#DADADA"),
    panel.grid.minor = element_blank()
  )
```


::: {.callout-note title="Summary: K-Nearest Neighbors (KNN) Classification"}

For this part, I built and tested the KNN algorithm from scratch using a custom dataset with a curved, non-linear decision boundary—great for visualizing how KNN handles complexity.
🔧 Dataset Creation

I generated synthetic training and test datasets with two features: x1 and x2. The target y was based on whether x2 was above the curve sin(4x1) + x1. So we’re not dealing with a simple straight-line split—this setup made it perfect for testing how well KNN adapts to non-linearity.
🛠️ Manual KNN Logic

Instead of using a library, I wrote my own KNN:

    Measured Euclidean distance from each test point to every training point

    Picked the k closest ones

    Took a majority vote from their labels to predict

Then, I ran class::knn() from R and matched it against my version—results were identical, so I knew the logic and execution were solid.
📈 Accuracy vs. K Plot

To figure out the best k, I looped through values from 1 to 30 and calculated accuracy on the test set for each. I plotted it out and saw a clear peak—some values of k perform way better than others, especially with non-linear boundaries.
🧠 Takeaways

    Smaller k (like 1–3): Super sensitive to noise. It overfits—memorizing every tiny bump.

    Larger k (like 25–30): Too smooth. It underfits and misses the real pattern.

    There’s a sweet spot right in the middle that nails the boundary without chasing noise.

This whole thing walked through a full pipeline—custom data, scratch implementation, matching it against the built-in function, then tuning k with visual insight. It was a cool way to not just use KNN, but actually understand why and when it works best.
:::









