---
title: "Multinomial Logit Model"
author: "Srujith Kapuluri"
date: 05-28-2025
---


This assignment expores two methods for estimating the MNL model: (1) via Maximum Likelihood, and (2) via a Bayesian approach using a Metropolis-Hastings MCMC algorithm. 


## 1. Likelihood for the Multi-nomial Logit (MNL) Model

Suppose we have $i=1,\ldots,n$ consumers who each select exactly one product $j$ from a set of $J$ products. The outcome variable is the identity of the product chosen $y_i \in \{1, \ldots, J\}$ or equivalently a vector of $J-1$ zeros and $1$ one, where the $1$ indicates the selected product. For example, if the third product was chosen out of 3 products, then either $y=3$ or $y=(0,0,1)$ depending on how we want to represent it. Suppose also that we have a vector of data on each product $x_j$ (eg, brand, price, etc.). 

We model the consumer's decision as the selection of the product that provides the most utility, and we'll specify the utility function as a linear function of the product characteristics:

$$ U_{ij} = x_j'\beta + \epsilon_{ij} $$

where $\epsilon_{ij}$ is an i.i.d. extreme value error term. 

The choice of the i.i.d. extreme value error term leads to a closed-form expression for the probability that consumer $i$ chooses product $j$:

$$ \mathbb{P}_i(j) = \frac{e^{x_j'\beta}}{\sum_{k=1}^Je^{x_k'\beta}} $$

For example, if there are 3 products, the probability that consumer $i$ chooses product 3 is:

$$ \mathbb{P}_i(3) = \frac{e^{x_3'\beta}}{e^{x_1'\beta} + e^{x_2'\beta} + e^{x_3'\beta}} $$

A clever way to write the individual likelihood function for consumer $i$ is the product of the $J$ probabilities, each raised to the power of an indicator variable ($\delta_{ij}$) that indicates the chosen product:

$$ L_i(\beta) = \prod_{j=1}^J \mathbb{P}_i(j)^{\delta_{ij}} = \mathbb{P}_i(1)^{\delta_{i1}} \times \ldots \times \mathbb{P}_i(J)^{\delta_{iJ}}$$

Notice that if the consumer selected product $j=3$, then $\delta_{i3}=1$ while $\delta_{i1}=\delta_{i2}=0$ and the likelihood is:

$$ L_i(\beta) = \mathbb{P}_i(1)^0 \times \mathbb{P}_i(2)^0 \times \mathbb{P}_i(3)^1 = \mathbb{P}_i(3) = \frac{e^{x_3'\beta}}{\sum_{k=1}^3e^{x_k'\beta}} $$

The joint likelihood (across all consumers) is the product of the $n$ individual likelihoods:

$$ L_n(\beta) = \prod_{i=1}^n L_i(\beta) = \prod_{i=1}^n \prod_{j=1}^J \mathbb{P}_i(j)^{\delta_{ij}} $$

And the joint log-likelihood function is:

$$ \ell_n(\beta) = \sum_{i=1}^n \sum_{j=1}^J \delta_{ij} \log(\mathbb{P}_i(j)) $$



## 2. Simulate Conjoint Data

We will simulate data from a conjoint experiment about video content streaming services. We elect to simulate 100 respondents, each completing 10 choice tasks, where they choose from three alternatives per task. For simplicity, there is not a "no choice" option; each simulated respondent must select one of the 3 alternatives. 

Each alternative is a hypothetical streaming offer consistent of three attributes: (1) brand is either Netflix, Amazon Prime, or Hulu; (2) ads can either be part of the experience, or it can be ad-free, and (3) price per month ranges from \$4 to \$32 in increments of \$4.

The part-worths (ie, preference weights or beta parameters) for the attribute levels will be 1.0 for Netflix, 0.5 for Amazon Prime (with 0 for Hulu as the reference brand); -0.8 for included adverstisements (0 for ad-free); and -0.1*price so that utility to consumer $i$ for hypothethical streaming service $j$ is 

$$
u_{ij} = (1 \times Netflix_j) + (0.5 \times Prime_j) + (-0.8*Ads_j) - 0.1\times Price_j + \varepsilon_{ij}
$$

where the variables are binary indicators and $\varepsilon$ is Type 1 Extreme Value (ie, Gumble) distributed.

The following code provides the simulation of the conjoint data.

:::: {.callout-note collapse="true"}
```{r}
# set seed for reproducibility
set.seed(123)

# define attributes
brand <- c("N", "P", "H") # Netflix, Prime, Hulu
ad <- c("Yes", "No")
price <- seq(8, 32, by=4)

# generate all possible profiles
profiles <- expand.grid(
    brand = brand,
    ad = ad,
    price = price
)
m <- nrow(profiles)

# assign part-worth utilities (true parameters)
b_util <- c(N = 1.0, P = 0.5, H = 0)
a_util <- c(Yes = -0.8, No = 0.0)
p_util <- function(p) -0.1 * p

# number of respondents, choice tasks, and alternatives per task
n_peeps <- 100
n_tasks <- 10
n_alts <- 3

# function to simulate one respondent’s data
sim_one <- function(id) {
  
    datlist <- list()
    
    # loop over choice tasks
    for (t in 1:n_tasks) {
        
        # randomly sample 3 alts (better practice would be to use a design)
        dat <- cbind(resp=id, task=t, profiles[sample(m, size=n_alts), ])
        
        # compute deterministic portion of utility
        dat$v <- b_util[dat$brand] + a_util[dat$ad] + p_util(dat$price) |> round(10)
        
        # add Gumbel noise (Type I extreme value)
        dat$e <- -log(-log(runif(n_alts)))
        dat$u <- dat$v + dat$e
        
        # identify chosen alternative
        dat$choice <- as.integer(dat$u == max(dat$u))
        
        # store task
        datlist[[t]] <- dat
    }
    
    # combine all tasks for one respondent
    do.call(rbind, datlist)
}

# simulate data for all respondents
conjoint_data <- do.call(rbind, lapply(1:n_peeps, sim_one))

# remove values unobservable to the researcher
conjoint_data <- conjoint_data[ , c("resp", "task", "brand", "ad", "price", "choice")]

# clean up
rm(list=setdiff(ls(), "conjoint_data"))
```
::::

## 3. Preparing the Data for Estimation

The "hard part" of the MNL likelihood function is organizing the data, as we need to keep track of 3 dimensions (consumer $i$, covariate $k$, and product $j$) instead of the typical 2 dimensions for cross-sectional regression models (consumer $i$ and covariate $k$). The fact that each task for each respondent has the same number of alternatives (3) helps.  In addition, we need to convert the categorical variables for brand and ads into binary variables.
```{r}
# Load necessary packages (again, if session restarted)
library(tidyverse)
```
```{r}
# Convert brand and ad to dummy variables
conjoint_data <- read_csv("/Users/sruji/Desktop/marketing/Blog/Project_5/conjoint_data.csv")
mnl_data <- conjoint_data %>%
  mutate(
    brand_N = as.numeric(brand == "N"),
    brand_P = as.numeric(brand == "P"),
    ad_yes  = as.numeric(ad == "Yes")
  ) %>%
  # Select relevant columns for modeling
  select(resp, task, choice, brand_N, brand_P, ad_yes, price)
```
```{r}
mnl_data %>%
  count(resp, task) %>%
  pull(n) %>%
  table()
```

## 4. Estimation via Maximum Likelihood

```{r}
# Log-likelihood function for MNL
log_likelihood_mnl <- function(beta, data) {
  # Extract covariates
  X <- as.matrix(data[, c("brand_N", "brand_P", "ad_yes", "price")])
  
  # Compute X * beta (linear utility)
  V <- X %*% beta
  
  # Group by respondent-task
  data$V <- as.numeric(V)
  
  probs <- data %>%
    group_by(resp, task) %>%
    mutate(
      expV = exp(V),
      denom = sum(expV),
      prob = expV / denom
    ) %>%
    ungroup()
  
  # Return NEGATIVE log-likelihood
  -sum(log(probs$prob[probs$choice == 1]))
}
```
```{r}
# Initial guess
init_beta <- rep(0, 4)

# Run optimization
mnl_mle <- optim(
  par = init_beta,
  fn = log_likelihood_mnl,
  data = mnl_data,
  method = "BFGS",
  hessian = TRUE
)
```
```{r}
# Estimated coefficients
beta_hat <- mnl_mle$par
names(beta_hat) <- c("brand_N", "brand_P", "ad_yes", "price")

# Standard errors from Hessian
se <- sqrt(diag(solve(mnl_mle$hessian)))

# Create tidy summary table
mnl_results <- tibble(
  Variable = names(beta_hat),
  Estimate = beta_hat,
  StdError = se
)

print(mnl_results)
```
#### These results align closely with the true underlying part-worths used in the simulation:

    Consumers strongly prefer Netflix (baseline: Hulu),

    They are moderately favorable toward Amazon Prime,

    They dislike advertisements, and

    Higher prices reduce the likelihood of selection.

MLE successfully recovers the structure of preferences embedded in the conjoint design.



## 5. Estimation via Bayesian Methods
```{r}
log_prior <- function(beta) {
  dnorm(beta[1], 0, 5, log = TRUE) +  # brand_N
  dnorm(beta[2], 0, 5, log = TRUE) +  # brand_P
  dnorm(beta[3], 0, 5, log = TRUE) +  # ad_yes
  dnorm(beta[4], 0, 1, log = TRUE)    # price
}

log_posterior <- function(beta, data) {
  -log_likelihood_mnl(beta, data) + log_prior(beta)  # use negative log-lik because MLE used it that way
}
```
```{r}
# MH sampler function
mh_sampler <- function(log_post_fn, data, n_iter = 11000, init = rep(0, 4)) {
  chain <- matrix(NA, nrow = n_iter, ncol = 4)
  chain[1, ] <- init
  current_lp <- log_post_fn(init, data)
  
  for (i in 2:n_iter) {
    # Propose new beta
    proposal <- chain[i - 1, ] + c(
      rnorm(1, 0, sqrt(0.05)),
      rnorm(1, 0, sqrt(0.05)),
      rnorm(1, 0, sqrt(0.05)),
      rnorm(1, 0, sqrt(0.005))
    )
    
    # Calculate log posterior for proposal
    prop_lp <- log_post_fn(proposal, data)
    
    # Acceptance probability
    accept_prob <- exp(prop_lp - current_lp)
    
    if (runif(1) < accept_prob) {
      chain[i, ] <- proposal
      current_lp <- prop_lp
    } else {
      chain[i, ] <- chain[i - 1, ]
    }
  }
  
  chain
}
```
```{r}
# Run sampler
set.seed(42)
mh_chain <- mh_sampler(log_posterior, mnl_data, n_iter = 11000)

# Discard burn-in
mh_post <- mh_chain[1001:11000, ]

# Convert to tibble for plotting/summary
mh_post_df <- as_tibble(mh_post, .name_repair = "minimal")
names(mh_post_df) <- c("brand_N", "brand_P", "ad_yes", "price")
```
```{r}
# Posterior summaries: mean and 95% credible intervals
bayes_summary <- mh_post_df %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "value") %>%
  group_by(Variable) %>%
  summarise(
    Mean = mean(value),
    `2.5%` = quantile(value, 0.025),
    `97.5%` = quantile(value, 0.975),
    .groups = "drop"
  )

print(bayes_summary)
```
```{r}
# Plot for price beta
library(ggplot2)

# Trace plot
ggplot(mh_post_df, aes(x = 1:nrow(mh_post_df), y = price)) +
  geom_line(alpha = 0.5) +
  labs(title = "Trace Plot: Price Coefficient",
       x = "Iteration", y = "Price Beta")

# Posterior histogram
ggplot(mh_post_df, aes(x = price)) +
  geom_histogram(color = "black", fill = "skyblue", bins = 40) +
  labs(title = "Posterior Distribution: Price Coefficient",
       x = "Price Beta", y = "Frequency")
```

```{r}
# Add standard deviations
bayes_full <- mh_post_df %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "value") %>%
  group_by(Variable) %>%
  summarise(
    Mean = mean(value),
    SD = sd(value),
    `2.5%` = quantile(value, 0.025),
    `97.5%` = quantile(value, 0.975),
    .groups = "drop"
  )

print(bayes_full)
```
#### These posterior summaries closely align with our MLE estimates, providing strong reassurance in the model's stability and inference.

## 6. Discussion

If we didn't simulate the data ourselves, we would still be able to interpret the parameter estimates in meaningful, real-world terms:

    The estimate β_Netflix > β_Prime indicates that, all else equal, consumers have a stronger preference for Netflix over Amazon Prime. This likely reflects perceived content quality, brand familiarity, or cultural dominance.

    A negative β_price aligns with economic intuition: as price increases, the likelihood of a product being chosen decreases. This confirms that consumers are price-sensitive.

    The negative coefficient on ad_yes suggests that consumers dislike advertisements — services that include ads are less likely to be selected, holding other features constant.

Together, these results validate that the model is capturing real trade-offs that consumers make between brand, ad exposure, and price.
Modeling Preference Heterogeneity: Multilevel / Hierarchical MNL

The MNL model estimated here assumes that all respondents share the same preferences — a single vector of coefficients β. However, in real-world applications, consumer preferences are not homogeneous.

To simulate and estimate a multilevel (hierarchical) MNL model, we would need to:

    Allow for individual-level parameters, where each respondent has their own β_i drawn from a population distribution:
    βi∼N(μ,Σ)
    βi​∼N(μ,Σ)

    Simulate heterogeneity by assigning each respondent a unique β_i from that distribution when generating their choices.

    Estimate both levels: individual-specific preferences (β_i) and population-level hyperparameters (μ, Σ). This would typically require a Bayesian hierarchical model using methods like Gibbs sampling or Hamiltonian Monte Carlo (e.g., via rstan or brms in R).

This hierarchical structure allows for richer insights and more realistic predictions, particularly useful in marketing strategy, segmentation, and personalized recommendations.










